---
title: "Formal Problem Statement"
output: html_document
---
 
```{r load_packages, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(knitr)
```
 
  
#### Project Name: Prediction of Dengue Fever
#### Client: Departamento de Salud de Puerto Rico
#### Date: `r format(Sys.time(), '%d %B %Y')`
#### By: Gwen Rino

## Overview
***
This goal of this project is to increase efficiency in staffing San Juan's public health clinic for the treatment of dengue disease, thus saving the Departamento de Salud de Puerto Rico an average of $20,000 annually to invest in other public health projects.  
  
The approach will be to use U.S. National Oceanic and Atmospheric Administration (NOAA) weather data to create a model that predicts the number of weekly cases of dengue disease so that necessary staffing levels can be anticipated. This will reduce incidences of overstaffing while also meeting minimum staff-to-patient ratios without overdependence on costly per diem nurses.   
  
*Note that this is an imagined problem for the purposes of the assignment, and while I have grounded the project in factual information about dengue virus, details about the hiring practices and needs of the Puerto Rican Department of Health are completely fabricated.*

## Background  
***
  
**The big picture: dengue’s impact**  
  
Dengue is a mosquito-borne virus that causes both dengue fever and the more severe dengue hemorrhagic fever (DHF). Incidence of the virus varies with the season and weather, and global climate change seems to be affecting both the range and the rate of the disease.  
  
There are about 96 million clinical manifestations of dengue disease worldwide each year, and that number is growing, especially in the Americas. The virus that causes dengue can induce symptoms such as fever, headache, liver dysfunction, and muscle, joint, and abdominal pain. DHF, a less common but more severe dengue infection, is a hemorrhagic illness that is often fatal, especially in children. There is no medication or treatment for dengue disease beyond rest and fluids. However, an early clinical diagnosis and fluid replacement therapy can mitigate the severity of the illness and, in the case of DHF, save lives. 
  
According to a study in *The Lancet Infectious Diseases*, the annual global cost of dengue disease is nearly $9 billion in hospitalization, ambulatory treatment, and lost wages. The global cost of the disease, and the human suffering that cost represents, can be addressed by the smart and efficient use of public health resources.

References:  
http://www.who.int/mediacentre/factsheets/fs117/en/    
https://ehp.niehs.nih.gov/118-a117/  
http://www.thelancet.com/journals/laninf/article/PIIS1473-3099%2816%2900146-8/fulltext  
  
**The local picture: optimizing public health resources**  
    
The Departamento de Salud de Puerto Rico (health department of Puerto Rico) must operate within budgetary limits, allocating resources to have the biggest possible positive impact on public health. Dengue management costs include vector control (mosquito abatement), vaccine research, public education, and direct patient care. Among these, the need for direct patient care is the most variable because the local incidence of disease varies significantly from week to week, from no diagnoses at all to several hundred cases per week during periods of outbreak.   
  
The accepted minimum ratio of care givers to dengue patients (ambulatory and hospitalized) is 1:9. Dengue disease is more prevalent in summer and fall than in winter and spring. The current staffing plan, instituted after the deadly dengue season of 2005, places 1 dengue specialist at the San Juan clinic in the spring, increases that number to 5 health care workers during the summer and fall, and reduces the number to 2 during the winter. The health department hires per diem nurses during outbreaks in order to meet the minimum required ratio.  

The plan helps control costs during years with outbreaks of disease, but leaves the clinic substantially overstaffed during quiet years, which is expensive. For example:

```{r, echo=FALSE}
table <- tibble(
  Year = c("2006, quiet year", "2007, outbreak year"),
  Overstaffed.wks = c("35", "6"),
  Overstaffing.cost = c("$32,500", "$3,500"),
  Understaffed.wks = c("2", "17"),
  Understaffing.cost = c("$1,200", "$10,200"),
  Total.extra.cost = c("$33,700", "$13,700")
)
kable(table, caption = "Staffing Costs from Over- and Understaffing, 2006-2007")
```


A model that predicts the number of weekly cases of dengue disease with adequate accuracy would allow public health administrators to optimize the allocation of resources for staffing, deploying health care workers when and where they are needed.    
  
It is worth noting that as global climate change shifts the range and incidence of dengue and other mosquito-borne diseases, public health resources will be increasingly stretched. Any model that helps makes decisions about the efficient and effective allocation of public health resources will be increasingly essential in the coming years and decades.  
  
Reference:  
https://nursesalaryguide.net/nurse-salary-in-puerto-rico/

## Key Roles
***
  
Model creation and deployment:  
-	Gwen Rino 
  
Model users:  
-	Departamento de Salud de Puerto Rico administration team  
  
Subject matter experts:  
-	Departamento de Salud de Puerto Rico dengue team  
-	Center for Disease Control, Dengue Branch  
-	U.S. National Oceanic and Atmospheric Administration  
  
Other stakeholders:  
-	Local health care workers  
-	Local dengue patients  

## Success Metrics
***
  
**Project success definition:**  
This project will be a success if the predicted number of weekly dengue cases allows the health department to staff the San Juan health clinic at the required 1:9 ratio (without overstaffing nor hiring per diem nurses) 75% of the time.  
  
**Required model performance:**  
  
Our initial goal for model performance is an Mean Absolute Error (MAE) <= 4.0. This level of performance is our estimate of what is required to achieve project success.

## Risk Estimates
***
  
**If the project success metric is not met:**  

There is an efficiency cost to the health department each time the clinic is over- or understaffed. The cost of overstaffing by one person is the weekly wage of a health care worker who wasn't needed ($500/week). The cost of understaffing by one person is the weekly wage of a per diem nurse ($600/week).

**If the model requirements are not met:**

If the model performance metric (MAE <= 4.0) is not met, then the model cannot predict the number of weekly dengue cases with adequate accuracy to confidently plan clinic staffing levels. The project goals will need to be revisited and revised.  

## Model Deployment
***

The model will be deployed as a tool that outputs predicted weekly cases of dengue disease and the coinciding recommended number of clinic staff to engage (that is, number of staff required to meet the acceptable staff-to-patient ratios for the predicted number of patients). 
  
This model deployment requires a new system for making staffing decisions. Health department administrators, who will be the users of the model, will need training in how to use the new system. Health care workers may be resistant to the new system if it requires them to be more flexible about their weekly work assignments than they are accustomed to.  
  

## Timeline
***

This project is undertaken in the context of the UC Extension course CompSciX415.1, which runs through May 30, 2018. An estimated timeline follows:

**Phase One**  
This Formal Problem Statement assignment addresses the Inception and Formalization phases of the project. All parties should be in agreement about the details of the Formal Problem Statement *by April 18.*

**Phase Two**  
The next phase is Model Development. After working on data management for a week, I expect to have preliminary results of the model *by May 2* and to have determined whether the model meets the performance requirements by *May 16.*

**Phase Three**  
Model Deployment will occupy the next two weeks and be complete *by May 23.*

**Phase Four**  
A plan for the final stage, Model Management, will be in place *by May 30.*

## Estimated cost of project
***

To be determined.