---
title: "Arima Model Performance"
author: "Gwen Rino"
date: "May 23, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, echo=FALSE, warning=FALSE, message=FALSE}
library(ProjectTemplate)
```

```{r echo=FALSE, results='hide', message=FALSE}
setwd("../../..")
load.project()
```
## Model Description  
  
This is a univariate time series model that utilizes the `auto.arima()` function to determine that the best fit ARIMA is an ARIMA(1,1,1) model.\
  
```{r}
# Fit model
arima.mod <- auto.arima(ts.selected[ ,"total_cases"])
summary(arima.mod)

```

## Model Evaluation  
  
I used the `forecast::tsCV()` function to cross validate this model using the forecast evaluation on a rolling origin method. I forecast at three horizons: 1 week ahead, 6 weeks ahead, and 6 months ahead. As expected, the MAE increases with a longer horizon forecast, from 7.9 for a 1 week horizon to 29.1 for a 6 month horizon.\
  
```{r}

# Function that creates forecast object
far <- function(x, h){forecast(Arima(x, order=c(1,1,1)), h=h)}

# 1 week horizon
e <- tsCV(ts.selected[ ,"total_cases"], far, h=1)
# Calculate MAE
print(paste("1 week horizon MAE = ", mean(abs(e), na.rm = TRUE)))

# 6 week horizon
e1 <- tsCV(ts.selected[ ,"total_cases"], far, h=6)
# Calculate MAE
print(paste("6 week horizon MAE = ", mean(abs(e1), na.rm = TRUE)))


# 6 month horizon
e2 <- tsCV(ts.selected[ ,"total_cases"], far, h=26)
# Calculate MAE
print(paste("6 month horizon MAE = ", mean(abs(e2), na.rm = TRUE)))

```



