---
title: "Dynamic Regression Model Performance"
author: "Gwen Rino"
date: "May 23, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, echo=FALSE, warning=FALSE, message=FALSE}
library(ProjectTemplate)
```

```{r echo=FALSE, results='hide', message=FALSE}
setwd("../../..")
load.project()
```

## Model Description  
  
To experiment with Dynamic Regression models, I first needed to select the most likely variables for use as the regressor in the xreg argument. I selected five features by looking for the intersection of the most likely variables identified by their `randomForest::importance()` according to a Random Forest model, and those identified by their `earth::evimp()` according to a Multivariate Adaptive Regression Spline (see "dengue/src/FeatureSelection/FeatureSelection.R"). These variables are:  
  
- nonres_guests
- station_max_temp_c
- reanalysis_tdtr_k
- reanalysis_dew_point_temp_k
- reanalysis_specific_humidity_g_per_kg  

Experimentation with these five variables revealed that an ARIMA(1,1,1) model with `reanalysis_dew_point_temp_k` as the regressor was the best (see "dengue/src/Models/4.0.DynamicRegression.R"). The model takes the form:\

```{r}

# Fit model
model <- auto.arima(ts.final[,"total_cases"], 
                    xreg = ts.final[,"reanalysis_dew_point_temp_k"])
summary(model)
```
  
In order to forecast `total_cases` with this model, we need a model for forecasting the xreg variable `reanalysis_dew_point_temp_k`. Experimentation revealed that a fine choice was a Seasonal Naive model (which uses the mean of the value from the season in the past--in this case, the season = the week of the year--as the predicted value for the season in the future). See "dengue/src/Models/4.DynamicRegression.R". This Seasonal Naive model for the xreg term is incorporated into the model for forecasting as below:\

```{r}

# Model of total_cases with reanalysis_dew_point_temp_k as regressor
model <- auto.arima(ts.final[,"total_cases"], 
                    xreg  = ts.final[,"reanalysis_dew_point_temp_k"])

# Model of dew point time series
dewpt.model <- snaive(ts.final[,"reanalysis_dew_point_temp_k"])

# Function to forecast using model (returns predictions)

DRfc <- function(h){
  # h is the forecast horizon in weeks
  ptval <- forecast(dewpt.model, h=h)[["mean"]] # predictions for use in xreg
  print(forecast(dengue.model, xreg = rep(ptval))[["mean"]]) # forecast, print results
}

```

## Model evaluation  
  
I used the `greybox::ro()` function to cross validate this model using the forecast evaluation on a rolling origin method (500 origins). I forecast at three horizons: 1 week ahead, 6 weeks ahead, and 6 months ahead. With an MAE of 1.1 for a 1 week horizon, 8.0 for 6 weeks ahead, and 18.0 for 6 months ahead, this model is substantially better than any previous one.\

```{r}

# Set up for cross validation
x <- ts.final[,"total_cases"]
xreg <- ts.final[,"reanalysis_dew_point_temp_k"]
ourCall <- "predict(arima(x=data, order=c(1,1,1), xreg=xreg[counti]), n.ahead=h, newxreg=xreg[counto])"
ourValue <- "pred"

## 1 week horizon
returnedValues1 <- ro(x,h=1,origins=500,ourCall,ourValue)
# Calculate MAE
print(paste("1 week horizon MAE = ", 
            mean(abs(returnedValues1$actuals -
                       returnedValues1$pred),na.rm = TRUE)))

## 6 week horizon
returnedValues6 <- ro(x,h=6,origins=500,ourCall,ourValue)
# Calculate MAE
print(paste("6 week horizon MAE = ", 
            mean(abs(returnedValues6$actuals -
                       returnedValues6$pred),na.rm = TRUE)))

## 6 month horizon
returnedValues26 <- ro(x,h=26,origins=500,ourCall,ourValue)
# Calculate MAE
print(paste("6 month horizon MAE = ", 
            mean(abs(returnedValues26$actuals -
                       returnedValues26$pred),na.rm = TRUE)))

```


