---
title: "Linear Model Performance"
author: "Gwen Rino"
date: "April 20, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Model description  

The linear model uses the features identified as having significant p-values in Pearson correlation tests against the target variable total_cases (see `dengue/src/EDA/EDA.R`). The features used in the model are:   
  
- `ndvi_nw`  
- `reanalysis_air_temp_k`  
- `reanalysis_avg_temp_k`     
- `reanalysis_dew_point_temp_k`  
- `reanalysis_max_air_temp_k`      
- `reanalysis_min_air_temp_k`            
- `reanalysis_relative_humidity_percent`    
- `reanalysis_specific_humidity_g_per_kg`  
- `station_avg_temp_c`  
- `station_max_temp_c`   
- `station_min_temp_c`  
- `weekofyear`    

I ran the model with missing values imputed two ways, first by median values and second by k nearest neighbor values. 

```{r}

source("~./src/LoadData/Load.R") # HELP
source("~./src/Munge/Munge.r") # HELP
source("~./src/Models/1.Linear.R") # HELP

# MEDIAN VALUE IMPUTATION

# Create training and test sets
set.seed(555)
train_set.1 <- dengue.med.sm %>% sample_frac(0.7)
test_set.1 <- anti_join(dengue.med.sm, train_set.1)

# Fit model
lm_1 <- lm(total_cases ~ ., data = train_set.1)

# KNN VALUE IMPUTATION

# Create training and test sets
set.seed(777)
train_set.2 <- dengue.knn.sm %>% sample_frac(0.7)
test_set.2 <- anti_join(dengue.knn.sm, train_set.2)

# Fit model
lm_2 <- lm(total_cases ~ ., data = train_set.2)
```

## Model evaluation 

Applying these models to a test set of a random 30% of the data, a comparison of the predicted to the actual number of cases yields a Mean Absolute Error (MAE) of 27.5 (median imputation model) and 26.4 (knn imputation model), both of which are actually worse than the Naive Model. It is noteworthy that by far the most influencial feature is `weekofyear`. I need to learn how to model and predict time series!

```{r}
# Predictions from lm_1 (MEDIAN VALUE IMPUTATION)
predictions.lm_1 <- predict(lm_1, newdata = test_set.1)
# Find MAE of error (actual number of cases - predicted number of cases)
mean(abs(test_set.1$total_cases - predictions.lm_1))


# Predictions from lm_2 (KNN VALUE IMPUTATION)
predictions.lm_2 <- predict(lm_2, newdata = test_set.2)
# Find MAE of error (actual number of cases - predicted number of cases)
mean(abs(test_set.2$total_cases - predictions.lm_2))

```






